```{r setup}
#| message: false
#| warning: false
#| include: false

library(tidyverse)
library(readxl)
list.files("r", pattern = "\\.R", full.names = TRUE) |> lapply(source)
```

## Open and prepare the raw dataset {.unnumbered}

A summary of the data from the terri4sol project is presented below. 

```{r}
data_t4s <- read.csv("data/soil_0.2.6.csv")
summary(data_t4s)
```


Soil organic carbon stocks were estimated as

$$ stock = C \cdot (\Delta d \cdot area) \cdot fine \cdot bd$$

where $C$ is the soil organic carbon content, $\Delta d$ is the measurement depth (0.1 m), $area$ is the standardisation area (1 ha = $10^4$ m), $fine$ is the fine fraction of the soil, and $bd$ is the bulk density of the fine fraction (in Mg m$^3$).

```{r}
data_t4s <- data_t4s |>
  mutate(stock = C / 100 * 1e3 * fine_fraction * bulk_density_g_cm3)
```

We also added data from the RévaTéné project that collected SOC data in the Téné forest plots. 

```{r}
data_tene <- read_excel("data/BdD_Mathus_TENE.xls", 2) |>
  mutate(T_stock = `CCOS_0-30cm` / 100 *
           (`Dapp_0-10cm` + `Dapp_10-20cm` + `Dapp_20-30cm`) / 3 *
           (1 - `%EG_0-30cm`) * 1e3) |>
  mutate(site = "tene", plot = SUBPLOT) |>
  separate(SUBPLOT, c("pl", "sub")) |>
  reframe(get_tene_coords(pl, sub), across()) |>
  select(site, plot, lon, lat, T_stock)
```

We then sum the carbon stocks over all depths at each site to get the total carbon stocks, and save the results.

```{r}
read.csv("data/plot_0.2.6.csv") |>
  # correct plot coordinates
  mutate(lonc = ifelse(lon > 0, lat, lon), latc = ifelse(lat < 0, lon, lat)) |>
  select(site, plot, lonc, latc) |>
  rename(lon = lonc, lat = latc) |>
  merge(data_t4s) |>
  group_by(site, plot, lon, lat) |>
  summarise(T_stock = sum(stock), .groups = "drop") |>
  bind_rows(data_tene) |>
  write.csv("data/cache/data_stocks.csv", row.names = FALSE)
```

