```{r setup}
#| message: false
#| warning: false

library(tidyverse)
```

## Open and prepare the raw dataset {.unnumbered}

```{r}
data_t4s <- read.csv("data/t4s_0.1.4.csv")
summary(data_t4s)
```

Soil organic carbon stocks were estimated as 

$$ stock = C \cdot (\Delta d \cdot area) \cdot fine \cdot bd$$ 

where $C$ is the soil organic carbon content, $\Delta d$ is the measurement depth (0.1 m), $area$ is the standardisation area (1 ha = $10^4$ m), $fine$ is the fine fraction of the soil, and $bd$ is the bulk density of the fine fraction (in Mg m$^3$).

```{r}
data_t4s <- data_t4s |> 
  mutate(stock = C/100 * 1e3 * fine_fraction * bulk_density_g_cm3)
```

We then sum the carbon stocks over all depths at each site to get the total carbon stocks, and save the results.

```{r}
data_stocks <- data_t4s |> 
  filter(!is.na(plot)) |> 
  group_by(site, plot) |> 
  summarise(ndepth = n(), T_stock = sum(stock))
data_plot <- data_t4s |> 
  filter(!is.na(plot)) |> 
  select(site, contains("plot"), contains("landuse"),  toposequence, lon, lat) |> 
  unique()

merge(data_stocks, data_plot, by = c("site", "plot")) |> 
  write.csv("data/data_T4s_final.csv", row.names = FALSE)
```


