```{r setup}
#| message: false
#| warning: false

library(tidyverse)
```

## Open and prepare the raw dataset {.unnumbered}

```{r}
data_t4s <- read.csv("data/soil_0.2.4.csv") 
summary(data_t4s)
```

Soil organic carbon stocks were estimated as

$$ stock = C \cdot (\Delta d \cdot area) \cdot fine \cdot bd$$

where $C$ is the soil organic carbon content, $\Delta d$ is the measurement depth (0.1 m), $area$ is the standardisation area (1 ha = $10^4$ m), $fine$ is the fine fraction of the soil, and $bd$ is the bulk density of the fine fraction (in Mg m$^3$).

```{r}
data_t4s <- data_t4s |> 
  mutate(stock = C/100 * 1e3 * fine_fraction * bulk_density_g_cm3)
```

We then sum the carbon stocks over all depths at each site to get the total carbon stocks, and save the results.

```{r}
read.csv("data/plot_0.2.4.csv") |> 
  select(site, plot, lon, lat) |> 
  merge(data_t4s) |> 
  group_by(site, plot, lon, lat) |> 
  summarise(T_stock = sum(stock), .groups = "drop") |> 
  write.csv("data/cache/data_stocks.csv", row.names = FALSE)
```
