```{r setup}
#| message: false
#| warning: false

library(terra)
library(gdalUtilities)
library(tidyverse)
library(ggcorrplot)
library(ggfortify)
```

## Soilgrids data {.unnumbered}

The Soilgrids data was retrieved from the [ISRIC website](https://soilgrids.org/), using SoilGrids 2.0 [@Poggio2021]. Variables codes are described [here](https://www.isric.org/explore/soilgrids/faq-soilgrids#What_do_the_filename_codes_mean).

```{r download-soilgrids}
# function to download and crop CHELSA data
dl_soilgrids <- function(
    var = "cec",
    depth = "0-5cm", 
    stat = "mean", 
    dir_local = "data/ISRIC",
    box = c(-9, -2, 4, 11)
) {
  
  # project bounding box into IGH
  ext_vect <- terra::as.polygons(terra::ext(box), crs = "EPSG:4326")
  igh <- '+proj=igh +lat_0=0 +lon_0=0 +datum=WGS84 +units=m +no_defs'
  ext_igh <- terra::ext(terra::project(ext_vect, igh))
  bb <- c(ext_igh$xmin, ext_igh$ymax, ext_igh$xmax, ext_igh$ymin)
  
  # create url and file name
  sg_url <- "/vsicurl/https://files.isric.org/soilgrids/latest/data/"
  datos <- paste0(var, "/", var, "_", depth, "_", stat, ".vrt")
  file <- paste0("data/ISRIC/", var, "_", depth, "_", stat, ".tif")
  
  # download from ISRIC site
  if (!file.exists(file)) {
    gdalUtilities::gdal_translate(
      paste0(sg_url,datos), 
      file ,
      tr=c(250,250),
      projwin=bb,
      projwin_srs =igh
    )
  }
}

if (! dir.exists("data/ISRIC")) dir.create("data/ISRIC")

for (v in c("bdod", "cec", "cfvo", "clay", "nitrogen", "phh2o", "sand", "silt")) {
  dl_soilgrids(
    var = v,
    depth = "0-5cm", 
    stat = "mean", 
    dir_local = "data/ISRIC",
    box = c(-9, -2, 4, 11)
  )
}
```

The values of SoilGrids variables (for now, at 0-5 cm depth) are then extracted at the location of all terri4sol plots.

```{r extract-soilgrids}
#| results: 'hide'

# extract values for all t4s plots
data_stocks <- read.csv("data/data_T4s_final.csv")

soilgrids_data <- 
  list.files("data/ISRIC", pattern = "tif$", full.names = TRUE) |> 
  rast() |> 
  project("EPSG: 4326") |> 
  terra::extract(data_stocks[, c("lon", "lat")]) |> 
  rename_with( ~ gsub(pattern = "_mean", replacement = "", x = .x)) |> 
  select(contains("cm")) |> 
  bind_cols(data_stocks[, c("site", "plot")])|> 
  relocate(site, plot)

write.csv(soilgrids_data, file = "data/soilgrids_data.csv", row.names = FALSE)
```

The following plots have NAs in at least one Soilgrids variable.

```{r}
soilgrids_data |> 
  filter(if_any(contains("cm"), is.na)) |> 
  relocate(site, plot)
```

Below is the result of a Principal Component Analysis with all Soilgrids variables.

```{r}
data_pca <- soilgrids_data |> 
  filter(!if_any(everything(), is.na)) 

data_pca |> 
  select(contains("cm")) |> 
  prcomp(scale = TRUE) |> 
  autoplot(loadings = TRUE, loadings.label = TRUE, 
           data = data_pca, colour = "site") +
  theme_minimal()
```

We can also plot the pairwise correlation of all Soilgrids variables.

```{r}
soilgrids_data |>  
  select(contains("cm")) |> 
  filter(!if_any(everything(), is.na)) |> 
  cor() |> 
  ggcorrplot()
```
