```{r setup}
#| message: false
#| warning: false
library(terra)
library(tidyverse)
```

# Topography {.unnumbered}

Elevation, slope, and aspect were derived from the Shuttle Radar Topography Mission (SRTM v3) digital elevation model (USGS SRTMGL1_003; 30 m resolution). Slope and aspect were computed from the original elevation data in Google Earth Engine, after which all bands were resampled to 100 m.

Topographical Wetness Index (TWI) was calculated using the same SRTM source. First, the 30 m DEM was aggregated to 100 m using a mean reducer. Sinks in the resampled DEM were filled using the Fill Sinks [@Wang2006] algorithm implemented in the SAGA GIS library (minimum slope = 0°). The SAGA Wetness Index module was then used to compute TWI (parameters: Suction = 10; Area = Specific Catchment Area; Slope type = Catchment slope; Minimum slope = 0; Offset slope = 0.1; Slope weighting = 1).

```{r}
#| message: false
#| warning: false
topo <- c(
  "data/01_Data/Environmental_variables/SRTM_V3_stack_100m_CI.tif",
  "data/01_Data/Environmental_variables/TWI_100m_CI.tif"
) |> rast()
names(topo)[4] <- "twi"
plot(topo)
```

```{r extract-thermal}
#| message: false
#| warning: false
# extract values for all t4s plots
coord_stocks <- read.csv("data/cache/data_stocks.csv") |>
  # project into EPSG:32630
  vect(crs = "epsg:4326") |>
  terra::project("epsg:32630")
topo |>
  terra::extract(coord_stocks) |>
  select(-ID) |>
  bind_cols(select(as.data.frame(coord_stocks), "site", "plot")) |>
  relocate(site, plot) |>
  write.csv(file = "data/cache/topo_data.csv", row.names = FALSE)
```

## Distribution of values in the t4s dataset

```{r}
#| message: false
#| warning: false
read.csv("data/cache/topo_data.csv") |>
  pivot_longer(cols = c("elevation", "slope", "aspect", "twi")) |>
  ggplot(aes(x = value)) +
  geom_density() +
  facet_wrap(~name, scales = "free") +
  theme_classic() +
  labs(x = NULL)
```

```{r}
#| message: false
#| warning: false
read.csv("data/cache/topo_data.csv") |>
  ggplot(aes(x = elevation, slope, col = site)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  theme_classic()
```

```{r}
#| message: false
#| warning: false
read.csv("data/cache/topo_data.csv") |>
  left_join(read.csv("data/plot_0.2.6.csv")) |>
  filter(!is.na(toposequence) & toposequence != "") |>
  ggplot(aes(x = toposequence, y = slope)) +
  geom_violin(fill = "grey") +
  theme_classic()
```
