```{r setup}
#| message: false
#| warning: false
library(terra)
library(sf)
library(tidyverse)
library(tidyterra)
```

## TMF data {.unnumbered}

We used data from the Tropical Moist Forests (TMF) product, which is based on Landsat images [@Vancutsem2021], to estimate the time since deforestation and the age of secondary forests. These classify tropical areas annually as follows: (i) undisturbed forests, (ii) disturbed forests, (iii) deforested areas, (iv) regrowing forests, (v) water, and (vi) other.

We estimated the age of secondary forests in 2022 as the number of years a pixel of regrowing forest in 2022 had continuously been classified as such. We estimated the time since deforestation as the time since a pixel that was classified as undisturbed or disturbed forest in 1990 (i.e. the first year of the TMF time series) had changed to another class.

```{r}
#| message: false
#| warning: false
#| eval: false

# South Cote d'Ivoire ("zone forestiere") area
area <- st_read("data/rci_south.shp") |> st_transform(4326)
# download tmf annual change for Cote d'Ivoire
if (!dir.exists("data/download/TMF")) dir.create("data/download/TMF")
for (yr in 1990:2024) {
  file <- paste0(
    "data/download/TMF/JRC_TMF_AnnualChange_v1_", yr,
    "_AFR_ID52_N10_W10.tif"
  )
  if (!file.exists(file)) {
    paste0(
      "https://ies-ows.jrc.ec.europa.eu/iforce/tmf_v1/download.py?",
      "type=tile&dataset=AnnualChange_", yr, "&lat=N10&lon=W10"
    ) |>
      download.file(file, mode = "wb")
  }
  crop_file <-
    paste0("data/download/TMF/JRC_TMF_AnnualChange_v1_", yr, "_RCI.tif")
  if (!file.exists(crop_file)) {
    rast(file) |>
      crop(ext(area)) |>
      writeRaster(crop_file)
  }
}
tmf <- list.files("data/download/TMF", "RCI", full.names = TRUE) |>
  grep(pattern = paste(seq(1990, 2022), collapse = "|"), value = TRUE) |>
  rast()
if (!dir.exists("data/cache/tiles")) dir.create("data/cache/tiles")
# create 25 smaller raster stacks
makeTiles(tmf, 2000, filename = "data/cache/tiles/tmf_.tif")

for (fl in list.files("data/cache/tiles", "tmf", full.names = TRUE)) {
  if (!file.exists(gsub("tmf", "tmf_sec", fl)) | # nolint
    !file.exists(gsub("tmf", "tmf_def", fl))) { # nolint
    r <- rast(fl)
    vals <- values(r)
  }
  # age of secondary forests in 2022
  if (!file.exists(gsub("tmf", "tmf_sec", fl))) {
    ind <- which(vals[, 33] == 4)
    age_sec <- vals[ind, ] |>
      apply(1, function(x) {
        if (any(!is.na(x) & x != 4)) {
          length(x) - max(which(x != 4))
        } else {
          NA
        }
      })
    vals_sec <- rep(NA, nrow(vals))
    vals_sec[ind] <- age_sec
    r_sec <- rast(
      matrix(vals_sec, nrow = nrow(r), ncol = ncol(r), byrow = TRUE),
      extent = ext(r), crs = crs(r)
    )
    writeRaster(r_sec, gsub("tmf", "tmf_sec", fl))
  }
  # deforestation age in 2022
  if (!file.exists(gsub("tmf", "tmf_def", fl))) {
    ind <- which(vals[, 1] %in% c(1, 2))
    age_def <- vals[ind, ] |>
      apply(1, function(x) {
        if (any(!is.na(x) & x > 2)) {
          sum(!is.na(x) & x > 2)
        } else {
          NA
        }
      })
    vals_def <- rep(NA, nrow(vals))
    vals_def[ind] <- age_def
    r_def <- rast(
      matrix(vals_def, nrow = nrow(r), ncol = ncol(r), byrow = TRUE),
      extent = ext(r), crs = crs(r)
    )
    writeRaster(r_def, gsub("tmf", "tmf_def", fl))
  }
}
list.files("data/cache/tiles", "sec", full.names = TRUE) |>
  lapply(rast) |>
  sprc() |>
  mosaic() |>
  writeRaster("data/cache/tmf_sec_age.tif", overwrite = TRUE)
list.files("data/cache/tiles", "def", full.names = TRUE) |>
  lapply(rast) |>
  sprc() |>
  mosaic() |>
  writeRaster("data/cache/tmf_def_age.tif", overwrite = TRUE)
# deforestation history in the region
def_age <- rast("data/cache/tmf_def_age.tif")
# undisturbed/degraded forest: deforestation age = 0
type <-
  "data/download/TMF/JRC_TMF_AnnualChange_v1_2022_RCI.tif" |>
  rast() |>
  crop(def_age)
weights <- focalMat(rast(ncols = 180, nrows = 180, xmin = 0), 10, "Gauss")
r0 <- ifel(type %in% c(0, 1), 0, def_age)
path <- "data/cache/tiles/"
if (!dir.exists(path)) dir.create(path)
makeTiles(r0, 2000,
  filename = paste0(path, "r0_.tif"),
  buffer = (nrow(weights) - 1) / 2
)

for (fl in list.files(path, "r0")) {
  n <- as.numeric(gsub("r0_|.tif", "", fl))
  i <- ceiling(n / 12)
  j <- n - 12 * floor((n - 1) / 12)
  x0 <- -8.561573
  y0 <- 8.513334
  ext0 <- c(x0 + 0.5389892 * c(j - 1, j), y0 - 0.5389892 * c(i, i - 1))
  paste0(path, fl) |>
    rast() |>
    focal(weights, fun = mean, na.rm = TRUE) |>
    crop(ext0) |> # remove buffer from tiles
    writeRaster(paste0(path, "smooth_", fl), overwrite = TRUE)
}
# mosaic, project into UTM and resample raster
rref <- "data/01_Data/Environmental_variables/TWI_100m_CI.tif" |>
  rast()
list.files(path, "smooth", full.names = TRUE) |>
  lapply(rast) |>
  sprc() |>
  mosaic() |>
  project(crs(rref)) |>
  resample(rref) |>
  writeRaster("data/cache/def_age_smooth.tif", overwrite = TRUE)
# remove tiles to save space
list.files("data/cache/tiles", full.names = TRUE) |> file.remove()
```

```{r include=FALSE}
#| message: false
#| warning: false
#| results: hide
area <- st_read("data/rci_south.shp") |> st_transform(4326)
```

```{r}
#| message: false
#| warning: false
ggplot() +
  geom_spatraster(data = rast("data/cache/tmf_sec_age.tif")) +
  scale_fill_viridis_c(na.value = "transparent") +
  geom_sf(data = area, fill = NA, col = "black") +
  labs(title = "Age of secondary forests in 2022", fill = "Age (yr)") +
  theme_minimal()
```

```{r}
#| message: false
#| warning: false
ggplot() +
  geom_spatraster(data = rast("data/cache/tmf_def_age.tif")) +
  scale_fill_viridis_c(na.value = "transparent") +
  geom_sf(data = area, fill = NA, col = "black") +
  labs(title = "Time since deforestation", fill = "Time (yr)") +
  theme_minimal()
```

We smoothed the time since deforestation (setting time = 0 for remaining forests) by averaging it with a 2D Gaussian kernel moving window. The standard deviation of the Guassian kernel was set to 300 m.

```{r}
#| message: false
#| warning: false
# extract values for all t4s plots
coord_stocks <- read.csv("data/cache/data_stocks.csv") |>
  # project into EPSG:32630
  vect(crs = "epsg:4326") |>
  terra::project("epsg:32630")

"data/cache/def_age_smooth.tif" |>
  rast() |>
  terra::extract(coord_stocks) |>
  rename(def_age = focal_mean) |>
  select(def_age) |>
  bind_cols(select(as.data.frame(coord_stocks), "site", "plot")) |>
  relocate(site, plot) |>
  write.csv(file = "data/cache/tmf_data.csv", row.names = FALSE)
```


```{r}
#| message: false
#| warning: false
data_site <- read.csv("data/cache/tmf_data.csv") |>
  bind_cols(select(geom(coord_stocks, df = TRUE), "x", "y")) |>
  filter(!is.na(x))
ggplot() +
  geom_spatraster(data = rast("data/cache/def_age_smooth.tif")) +
  scale_fill_viridis_c(na.value = "transparent") +
  geom_sf(data = area, fill = NA, col = "black") +
  geom_point(data = data_site, aes(x, y, col = is.na(def_age))) +
  labs(
    title = "Smoothed time since deforestation", fill = "Time (yr)",
    col = "Site has\nnon NA value"
  ) +
  theme_minimal()
```

## Comparison of TMF and field-based plot classification

```{r}
#| message: false
#| warning: false
tmf2022 <- rast("data/download/TMF/JRC_TMF_AnnualChange_v1_2022_RCI.tif") |>
  terra::extract(read.csv("data/cache/data_stocks.csv")[, c("lon", "lat")]) |>
  mutate(tmf_class = c(
    "Undisturbed", "Degraded", "Deforested", "Regrowth", "Water", "Other"
  )[Dec2022]) |>
  bind_cols(read.csv("data/cache/data_stocks.csv")) |>
  select(site, plot, tmf_class)
read.csv("data/cache/data_landuse.csv") |>
  inner_join(tmf2022) |>
  with(table(tmf_class, present_landuse))
```
