```{r setup}
#| message: false
#| warning: false

library(tidyverse)
library(usdm)
library(knitr)
```

## Variable selection {.unnumbered}

### Based on Variance Inflation Factor (VIF)

The `vifstep` function in R detects and removes multicollinearity between predictor variables by calculating the variance inflation factor (VIF). Multicollinearity occurs when predictor variables are highly correlated, which can lead to problems in regression models.

Using `vifstep` on a dataset, with a defined threshold (e.g. th=10), the function calculates the VIF for each variable. If a variable exceeds this threshold, it is deemed too correlated with the others and is removed from the dataset. This process is repetitive: after each deletion, the VIFs are recalculated for the remaining variables, until no variable has a VIF above the set threshold.

### Climate

After applying the VIF on the **temperature** variables, the selected variables that do not present a collinearity problem and that will be evaluated again for their relevance are :

```{r}
temp <- read.csv("data/data_T4s_final.csv") %>% 
  merge(read.csv("data/climate_data.csv")) %>% 
  select(site, plot, T_stock, paste0("bio", 1:11))

temp_vif <- temp |> select(contains("bio")) |> vifstep(th=10)
temp_vif@results |> kable()
```

After applying the VIF on the **precipitation** variables, the selected variables that do not present a collinearity problem and that will be evaluated again for their relevance are :

```{r}
prec <- read.csv("data/data_T4s_final.csv") %>% 
  merge(read.csv("data/climate_data.csv")) %>% 
  select(site, plot, T_stock, paste0("bio", 12:19))

prec_vif <- prec |> select(contains("bio")) |> vifstep(th=10)
prec_vif@results |> kable()
```

### Soil

After applying the VIF on the soil **physical properties**, the selected variables that do not present a collinearity problem and that will be evaluated again for their relevance are :

```{r}
phys <- read.csv("data/data_T4s_final.csv") %>% 
  merge(read.csv("data/soilgrids_data.csv")) %>%
  select(site, plot, T_stock, 
         paste0(c("bdod", "cfvo", "clay", "sand", "silt"), "_0.5cm"))

phys_vif <- phys |> select(-site, -plot, -T_stock) |> vifstep(th=10)
phys_vif@results |> kable()
```

After applying the VIF on the soil **chemical properties**, the selected variables that do not present a collinearity problem and that will be evaluated again for their relevance are :

```{r}
chem <- read.csv("data/data_T4s_final.csv") %>% 
  merge(read.csv("data/soilgrids_data.csv")) %>%
  select(site, plot, T_stock, 
         paste0(c("cec", "nitrogen", "phh2o"), "_0.5cm"))      

chem_vif <- chem |> select(-site, -plot, -T_stock) |> vifstep(th=10)
chem_vif@results |> kable()
```

### All variables

After applying the VIF all variables together, the selected variables that do not present a collinearity problem and that could be evaluated again for their relevance are :

```{r}
all <- read.csv("data/data_T4s_final.csv") %>% 
  merge(read.csv("data/climate_data.csv")) %>% 
  merge(read.csv("data/soilgrids_data.csv")) %>%
  select(site, plot, T_stock, contains("bio"), contains("cm"))      

all_vif <- all |> select(-site, -plot, -T_stock) |> vifstep(th=10)
all_vif@results |> kable()
```

## Based on Forward Feature Selection (FFS)

The Forward Feature Selection (FFS) method, applied via the `ffs` function, is a variable selection technique used in machine learning and statistics to identify the subsets of variables (features) most relevant to the construction of a predictive model.

How Forward Feature Selection (FFS) works:

-   **Initialization**: The method starts with a simple model, generally without any explanatory variables (or with only the constant in the case of regression).

-   **Progressive addition of variables**: At each iteration, the method examines the addition of each variable not yet selected to the current model. The variable that most improves model performance, according to a specific criterion (e.g. increasing R², reducing mean squared error, etc.), is added to the model.

-   **Validation** : Model performance is often validated at each stage using a cross-validation technique to avoid overlearning (overfitting).

-   **Stop criterion**: The process continues until there is no further significant improvement in model performance by adding new variables. A performance threshold can be predefined to determine when to stop adding variables.

-   **Final result**: At the end of the process, the model retains the variables that have been selected over the iterations as the most important for prediction.
